Analysis: Machine Learning - ReqAmt - Outliers

### 2021-12-12 code/analysis_code/analysis-ML-RA-outlier.Rmd
Last ML model update

It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis.

@@

#save file
DT_auto_fig_file = here("results","DT-auto-plot.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

@@

#save file
DT_det_fig_file = here("results","DT-det-plot.png")
ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                                     roundint = F,
                                                     type = 5,
                                                     digits = 5,
                                                     main = "Selected Decision Tree Model",
                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
                                                     shadow.col = "grey65",
                                                     col = "grey99")
#save manually as "DT-plot.png" -- ggsave throws an error

@@

#save file
DT_pred_act_file = here("results","DT-pred-act.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

@@

#save file
DT_resid_file = here("results","DT-resid.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

@@

#save file
DT_pred_resid_file = here("results","DT-pred-resid.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

@@

The identified best performing tree model has some interesting results. It identifies percent of state counties awarded IH program, New York, longitude, total number of agencies, FEMA cost share average, total response days, and an incident month of November for predicting requested FEMA funding. New York and November are particularly surprising key predictors. It is interesting that incident type or declaration type are not included as significant. The RMSE actually increases from the Null, which suggests this is not an appropriate model. Moving on.

@@

#save file
RF_auto_fig_file = here("results","RF-auto-plot.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

@@

#save manually as "RF-plot.png" -- ggsave throws errors

@@

#save file
RF_pred_act_file = here("results","RF-pred-act.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

@@

#save file
RF_resid_file = here("results","RF-resid.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

@@

#save file
RF_pred_resid_file = here("results","RF-pred-resid.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

@@

In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree. It identifies total number of federal agencies involved, percent of state counties awarded IH, FEMA cost share average, percent of counties awarded HM, and whether or not IH was awarded to any part of the state as the five most important predictors. Let's try another model to see if we can find an improvement in the RMSE.

@@

#save file
EN_auto_fig_file = here("results","EN-auto-plot.png")
ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
```

@@

#save manually as "EN-plot.png" -- ggsave throws errors

@@

#save file
EN_pred_act_file = here("results","EN-pred-act.png")
ggsave(filename = EN_pred_act_file, plot = EN_pred_act)

@@

#save file
EN_resid_file = here("results","EN-resid.png")
ggsave(filename = EN_resid_file, plot = EN_resids)

@@

#save file
EN_pred_resid_file = here("results","EN-pred-resid.png")
ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)

@@

#compare model performance to null model (and other models)
EN_RMSE <- EN_res %>%
              tune::show_best(n = 1) %>%
              dplyr::transmute(
                rmse = round(mean, 3),
                SE = round(std_err, 4),
                model = "EN") %>%
             dplyr::bind_rows(RF_RMSE)
EN_RMSE
```
The Elastic Net model places great emphasis on FEMA Cost Share average above all other predictors. The next closest predictors include Hawaii, New York, New Mexico, and DC as key predictors. The EN model has an improved RMSE, but the null model technically still has a lower RMSE.

@@

#save file
lasso_auto_fig_file = here("results","lasso-auto-plot.png")
ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
```

@@

#save manually as "LASSO-plot.png" -- ggsave throws errors

@@

#save file
lasso_pred_act_file = here("results","lasso-pred-act.png")
ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)

@@

#save file
lasso_resid_file = here("results","lasso-resid.png")
ggsave(filename = lasso_resid_file, plot = lasso_resids)

@@

#save file
lasso_pred_resid_file = here("results","lasso-pred-resid.png")
ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)

@@

## Model Selection and Evaluation
Based on the results of the models above, the null model actually has the best fit, which warrants further investigation (Especially with the outliers). In the mean time, we'll fit the LASSO model as it has the best fit out of the models.
```{r}
#fit model to training set but evaluate on the test data
lasso_fit_test <- best_lasso_wflow %>%
                    tune::last_fit(split = req_data_split)

#compare test performance against training performance
lasso_rmse_test <- collect_metrics(lasso_fit_test) %>%
                   dplyr::select(rmse = .estimate) %>%
                   dplyr::mutate(data = "test")

lasso_comp <- lasso_RMSE %>%
                  dplyr::filter(model == "LASSO") %>%
                  dplyr::transmute(rmse, data = "train") %>%
                  dplyr::bind_rows(lasso_rmse_test) %>%
                  dplyr::slice(-3) #don't know why the third row shows up
lasso_comp
#RMSEs are different but not as much as previous iterations

#save combined table
test_comp_file = here("results", "test-train-model-comp.Rds")
saveRDS(lasso_comp, file = test_comp_file)

#find predictions and intervals
lasso_resid_fit <- lasso_fit_test %>%
                      broom.mixed::augment() %>%
                      dplyr::select(.pred, log_req_amt) %>%
                      dplyr::mutate(.resid = .pred - log_req_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
lasso_pred_act_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = log_req_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point(color = "#3b528b") +
                        labs(title = "LASSO Test Fit: Predicted vs. Actual Log(Requested Funds)",
                              x = "Predicted Log(Requested Funds) ($)",
                              y = "Actual Log(Requested Funds) ($)")
lasso_pred_act_test

#save file
lasso_pred_act_test_file = here("results","lasso-pred-act-test.png")
ggsave(filename = lasso_pred_act_test_file, plot = lasso_pred_act_test)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
lasso_resids_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = as.numeric(row.names(lasso_resid_fit)), y = .resid))+
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
                      geom_point(color = "#3b528b") +
                      labs(title = "LASSO Test Fit: Residuals",
                            x = "Observation Number",
                            y = "Residual")
lasso_resids_test

#save file
lasso_resid_test_file = here("results","lasso-resid-test.png")
ggsave(filename = lasso_resid_test_file, plot = lasso_resids_test)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
lasso_pred_resid_test <- ggplot2::ggplot(lasso_resid_fit, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "LASSO Test Fit: Residuals vs Fitted Log(Requested Funds)",
                                x = "Predicted Log(Requested Funds) ($)",
                                y = "Residuals")
lasso_pred_resid_test

#save file
lasso_pred_resid_test_file = here("results","lasso-pred-resid-test.png")
ggsave(filename = lasso_pred_resid_test_file, plot = lasso_pred_resid_test)
```

>>>>

It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis where the log(ReqAmt) was less than zero.

@@

#save file
DT_auto_fig_file = here("results","DT-auto-plot-outlier.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

@@

#save file
DT_det_fig_file = here("results","DT-det-plot-outlier.png")
ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                                     roundint = F,
                                                     type = 5,
                                                     digits = 5,
                                                     main = "Selected Decision Tree Model",
                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
                                                     shadow.col = "grey65",
                                                     col = "grey99")
#save manually as "DT-plot-outlier.png" -- ggsave throws an error

@@

#save file
DT_pred_act_file = here("results","DT-pred-act-outlier.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

@@

#save file
DT_resid_file = here("results","DT-resid-outlier.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

@@

#save file
DT_pred_resid_file = here("results","DT-pred-resid-outlier.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

@@

The identified best performing tree model has some interesting results. It identifies percent of state counties awarded IH program, total number of agencies involved, total number of counties per state, percent of counties awarded PA program, FEMA Region VI, incident month of October and March as important predictors. It is interesting that incident type or declaration type are not included as significant. The RMSE does decrease from the null, which suggests the outlier removal was a necessary step. Let's explore the other models.

@@

#save file
RF_auto_fig_file = here("results","RF-auto-plot-outlier.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

@@

#save manually as "RF-plot-outlier.png" -- ggsave throws errors

@@

#save file
RF_pred_act_file = here("results","RF-pred-act-outlier.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

@@

#save file
RF_resid_file = here("results","RF-resid-outlier.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

@@

#save file
RF_pred_resid_file = here("results","RF-pred-resid-outlier.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

@@

In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree. It identifies percent of counties awarded IH program, total number of federal agencies involved, percent of counties awarded HM program, average FEMA cost share, and whether or not the IH program was awarded to any county in the state as the five most important predictors. Let's try another model to see if we can find an improvement in the RMSE.

@@

#save file
EN_auto_fig_file = here("results","EN-auto-plot-outlier.png")
ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
```

@@

#save manually as "EN-plot-outlier.png" -- ggsave throws errors

@@

#save file
EN_pred_act_file = here("results","EN-pred-act-outlier.png")
ggsave(filename = EN_pred_act_file, plot = EN_pred_act)

@@

#save file
EN_resid_file = here("results","EN-resid-outlier.png")
ggsave(filename = EN_resid_file, plot = EN_resids)

@@

#save file
EN_pred_resid_file = here("results","EN-pred-resid-outlier.png")
ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)

@@

#compare model performance to null model (and other models)
EN_RMSE <- EN_res %>%
              tune::show_best(n = 1) %>%
              dplyr::transmute(
                rmse = round(mean, 3),
                SE = round(std_err, 4),
                model = "EN") %>%
             dplyr::bind_rows(RF_RMSE)
EN_RMSE
```
The Elastic Net model places great emphasis on Maine being the state receving the declaration. The next closest predictors include average FEMA cost share, Vermont, Montana, and Idaho. The EN model has an improved RMSE from the DT model but not the RF model. Let's try LASSO just to round out the analysis.

@@

#save file
lasso_auto_fig_file = here("results","lasso-auto-plot-outlier.png")
ggsave(filename = lasso_auto_fig_file, plot = lasso_auto_plot)
```

@@

#save manually as "LASSO-plot-outlier.png" -- ggsave throws errors

@@

#save file
lasso_pred_act_file = here("results","lasso-pred-act-outlier.png")
ggsave(filename = lasso_pred_act_file, plot = lasso_pred_act)

@@

#save file
lasso_resid_file = here("results","lasso-resid-outlier.png")
ggsave(filename = lasso_resid_file, plot = lasso_resids)

@@

#save file
lasso_pred_resid_file = here("results","lasso-pred-resid-outlier.png")
ggsave(filename = lasso_pred_resid_file, plot = lasso_pred_resid)

@@

## Model Selection and Evaluation
Based on the results of the model comparison table, all of the models have a similar fit. However, RF has the lowest RMSE, so we will fit that to the test data.
```{r}
#fit model to training set but evaluate on the test data
RF_fit_test <- best_RF_wflow %>%
                    tune::last_fit(split = req_data_split)

#compare test performance against training performance
RF_rmse_test <- collect_metrics(RF_fit_test) %>%
                   dplyr::select(rmse = .estimate) %>%
                   dplyr::mutate(data = "test")

test_comp <- RF_RMSE %>%
                  dplyr::filter(model == "Random Forest") %>%
                  dplyr::transmute(rmse, data = "train") %>%
                  dplyr::bind_rows(RF_rmse_test) %>%
                  dplyr::slice(-3) #don't know why the third row shows up
test_comp
#RMSEs are very different (but lower in test data?)

#save combined table
test_comp_file = here("results", "test-train-model-comp-outlier.Rds")
saveRDS(test_comp, file = test_comp_file)

#find predictions and intervals
RF_resid_fit <- RF_fit_test %>%
                      broom.mixed::augment() %>%
                      dplyr::select(.pred, log_req_amt) %>%
                      dplyr::mutate(.resid = .pred - log_req_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
RF_pred_act_test <- ggplot2::ggplot(RF_resid_fit, aes(x = .pred, y = log_req_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point(color = "#3b528b") +
                        labs(title = "RF Test Fit: Predicted vs. Actual Log(Requested Funds)",
                              x = "Predicted Log(Requested Funds) ($)",
                              y = "Actual Log(Requested Funds) ($)")
RF_pred_act_test

#save file
RF_pred_act_test_file = here("results","RF-pred-act-test-outlier.png")
ggsave(filename = RF_pred_act_test_file, plot = RF_pred_act_test)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
RF_resids_test <- ggplot2::ggplot(RF_resid_fit, aes(x = as.numeric(row.names(RF_resid_fit)), y = .resid))+
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
                      geom_point(color = "#3b528b") +
                      labs(title = "RF Test Fit: Residuals",
                            x = "Observation Number",
                            y = "Residual")
RF_resids_test

#save file
RF_resid_test_file = here("results","RF-resid-test-outlier.png")
ggsave(filename = RF_resid_test_file, plot = RF_resids_test)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid_test <- ggplot2::ggplot(RF_resid_fit, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "RF Test Fit: Residuals vs Fitted Log(Requested Funds)",
                                x = "Predicted Log(Requested Funds) ($)",
                                y = "Residuals")
RF_pred_resid_test

#save file
RF_pred_resid_test_file = here("results","RF-pred-resid-test-outlier.png")
ggsave(filename = RF_pred_resid_test_file, plot = RF_pred_resid_test)
```
